Categorical and gradient aspects of wordlikeness\thanks{Thanks to Gene Buckley, Charles Yang, and audiences at NELS 43 for helpful comments.}

### 2013-01-28 3.tex
Revisions! Lots of 'em.

It has long been suggested that statistical criteria could distinguish between \emph{accidental phonotactic gaps}, those which could arise without any antecedent cause, and those gaps which are \emph{structural} in nature \citep[e.g.,][]{Fischer-Jorgensen1952,Saporta1955,Saporta1958,Vogt1954}.
Following seminal work by \citet{Mester1988} and \citet{McCarthy1988}, it is now commonly assumed that the phonotactic patterns in lexical entries ``directly determine the mental representation of the phonotactic constraints'' \citep[180]{Frisch2004} acquired by speakers, and therefore phonotactic constraints, whether categorical or gradient, can be directly inferred from statistical analysis of the lexicon.
There is by now an enormous amount of research that adopts this assumption.
Semitic languages, especially Arabic \citep{Coetzee2008a,Frisch2004,McCarthy1988,Pierrehumbert1993} but also Berber \citep{Elmedlaoui1995} and Tigrinya \citep{Buckley1997}, and Austronesian languages, including Javanese \citep{GraffInPress,Mester1988}, Muna \citep{Anttila2008a,Coetzee2008a}, and Samoan \citep{AldereteInPress}, have been of particular interest to linguists adopting this hypothesis; Aymara \citep{GraffInPress}, Cantonese \citep{Yip1989}, English \citep{Berkley1994a,Berkley1994b,Berkley2000,Coetzee2008b,Davis1989,Martin2007,Martin2011}, Gitksan \citep{Brown2010}, Hungarian \citep{Grimes2010}, Japanese \citep{Kawahara2006}, Ju|'hoansi \citep{Kinney2005}, Navajo \citep{Martin2007,Martin2011}, Ofo \citep[38f.]{MacEachern1999}, Russian \citep{Padgett1991,Padgett1992}, Shona \citep{Hayes2008a}, and Wargamay \citep{Hayes2008a} have also received a statistical treatment. 
\citet{Pozdniakov2007} analyze lexical tendencies in a diverse sample of 30 languages.

@@

It would be a result of great interest were it shown that statistical significance is both necessary and sufficient to identify linguistic generalizations which are internalized by speakers, but there is no reason to grant this assumption with respect to phonotactic knowledge.
Nothing demands that the antecedent cause of a statistically reliable lexical tendency be grammatical. 
There are reasons to suspect that many static phonotactic constraints identified in this manner have no synchronic reality.

@@

\noindent
However, subsequent sound changes \citep[e.g.,][]{Baldi1994,Safarewicz1932}, particularly the degemination of Old Latin \emph{ss} after diphthongs and long monophthongs (e.g., \emph{caussa} $>$ \emph{causa} `cause'), introduce numerous exceptions to \textsc{Rhotacism}.
In Classical Latin, intervocalic \emph{s} is found root-internally (\emph{asellus} `donkey', \emph{casa} `hut'), in environments derived by inflectional suffixes (\emph{uās}-\emph{uāsis} `vase', \emph{uisēre} `to view'), prefixation (\emph{dēsecāre} `to cut off'; cf.~\emph{dē} `from', \emph{secāre} `to cut'), compounding (\emph{olusātrum} `parsnip'; cf.~\emph{olus} `vegetable', \emph{ātrum} `black'), and denominal adjective formation (\emph{uentōsus} `windy'; cf.~\emph{uentus} `wind'), and is tolerated in nativized loanwords from Celtic (\emph{omāsum} `tripe'), Germanic (\emph{glaesum} `amber'), and Greek (\emph{basis} `pedestal').
These facts lead Saussure to conclude that \textsc{Rhotacism} is no longer ``inhérente à la nature de la langue'' (202).
While some of the apparent exceptions may be the result of opaque phonological application \citep{Heslin1987}---an explanation not yet available in Saussure's time---any formulation of rhotacization will admit nearly as many lexical exceptions as there are roots exhibiting \emph{s}-\emph{r} alternations \citep{GormanInPressc}.
Any synchronic account of the underrepresentation of intervocalic \emph{s} must confront the unproductive nature (as indicated by the accumulation of exceptions) of the proximate explanation for this tendency.

% FIXME mention Fudge's analysis of short-a before voiceless fricatives
% \citet{Trager1930,Trager1934,Trager1940},
% \citet{Cohen1970}
% \citet{Ferguson1975}
% \citet{PLC1} 429f.
% \citet{Fudge1987}
% \citet{AOE} I.203f., 515f. for an attempt to write a rule
% \citet{Payne1976,Payne1980} who finds that whereas speakers from out of state acquire 5 vocalic sound changes, only one out of 34 interviewees acquired the core pattern
% Fudge(!)

Other cases show that constraints identified by statistical analysis are inert: that is, there is no evidence (of any of type described in \S\ref{s:espt} above) that they are internalized by speakers.
For instance, most instances of Modern English [ʃ] derive from Old English [sk] (e.g., \emph{fisc} `fish') via sound change.
Since Old English does not permit long vowels before complex codas, compared to similar segments like [s], [ʃ] is still rarely preceded by long vowels in word-final syllables.\footnote{
    This sample is drawn from the CMU pronunciation dictionary, filtered by excluding words with a token frequency of less than 1 per million words in the SUBTLEX-US frequency norms.
    Less restrictive samples give similar results.
    The 9 words ending with a long vowel-[ʃ] sequence are \emph{douche}, \emph{leash}, \emph{gosh}, \emph{josh}, \emph{posh}, \emph{squash}, \emph{unleash}, \emph{wash}, and \emph{woosh}.}
As can be seen, long vowels are twice as common before [s] as before [ʃ], a significant generalization according to the Fisher exact test.
Similarly, \citet{HayesInPress} report that this constraint is discovered by the \citet{Hayes2008a} phonotactic learning model, which uses a related statistical criterion to identify constraints.
Despite this, \citet{Iverson2005} label the constraint on long vowels before [ʃ] as ``phonologically accidental'', as a millennium of coinages (e.g., \emph{posh}) and loanwords (e.g., \emph{douche}) disregard this generalization.
\citet{HayesInPress} find that a variant of this restriction has little to no effect on wordlikeness judgements.

\begin{table}[t]
\centering
\begin{tabular}{l r r r r}
\toprule
          & \{ɪ, ɛ, æ, ʌ, ʊ\}\gap{}\# & \{i, e, ɑ, ɔ, u\}\gap{}\# & \% long & $p$-value \\
\midrule
\gap{}ʃ\# & 78                & 9                 & 8\%      & \multirow{2}{*}{.026} \\
\gap{}s\# & 410               & 107               & 16\%     & \\
\bottomrule
\end{tabular}
\caption{Type frequencies of short and long vowels before word-final [ʃ] and [s]}
\label{ssh}
\end{table}

This latter example makes it clear that a purely statistical criterion overgenerates in the sense that it predicts phonotactic constraints which speakers do not seem to internalize.
To account for the cases like the one above, \citet{HayesInPress} propose that speakers are biased in favor of ``natural generalizations'' in probabilistic phonotactic learning.\footnote{
    \citeauthor{HayesInPress} do not provide an operational definition of ``naturalness'', so it is difficult to evaluate their specific results or to determine whether they apply the classification of ``natural'' and ''unnatural'' is applied in an ad hoc (or post hoc) fashion.}
However, this chapter argues that this overgeneration gives the lie to the broader assumption that phonotactics are extracted from patterns in the lexicon.
Rather, the only restrictions which speakers internalize are those which derive from phonological processes in the language.
This is independent of ``naturalness'', since both ``natural'' and ``unnatural'' variants of statistically reliable static constraints are shown to be inert.

@@

\citet{Lees1966b,Lees1966a} proposes three constraints on Turkish vowel sequences; these constraints are the focus of many subsequent studies. In this section, these constraints are formalized, and where possible, related to phonological alternations and to behavioral evidence bearing on speakers' knowlege of the restrictions. The following feature specification for the eight vowels of Turkish is assumed throughout.

@@

Two less familiar notations are used in this chapter. First, directional application conditions are assumed, so as to derive the left-to-right, iterative properties of the harmony rules. Since it was first proposed by \citet{Johnson1972}, considerable evidence for directional application has been adduced (e.g., \citealt{A74}: chap.~9, \citealt{GormanInPressc}, \citealt{Howard1972}:65f., \citealt{Kavitskaya2008}, \citealt{Kaye1982}, \citealt{KK77}:189f., \citeyear{KK79}:319f., \citealt{Piggott1975}, \citealt{Sohn1971}; see \citealt{McCarthy2003b} and \citealt{Wolf2011b} for recent reviews.) Further, \citet{Johnson1972} and \citet{Kaplan1994} prove directional application and simultaneous application are equivalent in terms of formal learnability.

Secondly, rather than the use of an unbounded number of Greek-letter variables ($\alpha$, $\beta$, etc.) over feature values $\{+, -\}$, only a single variable, denoted by `$=$', is used \citep{McCawley1973}. A structural description [$=$F]\ldots{}[$=$F] matches a string S$_i$\ldots{}S$_j$ if and only if S$_i$ and S$_j$ are both [$+$F] or both [$-$F]. This is more restrictive than Greek-letter notation, in that it prevents the value of one feature being applied to different feature; \citet{Odden2012} argues that the two counterexamples against this restriction presented in \emph{SPE} (352-353) are not probative.

@@

\citet[212]{A74} and \citet{Iverson1978} argue that suffix harmony in disharmonic roots found in (\ref{turkexcept}a) requires the rule governing suffix harmony alternations to be distinguished from a sequence structure constraint governing root harmony. 
Since the rule and sequence structure constraint are otherwise identical, this constitutes a ``duplication''. 
However, the theory of exceptionality proposed in \emph{SPE} can account for these facts without duplication \citep[197f.]{Zonneveld1978}.\footnote{
    \citet[29f.]{Kiparsky1968} discusses a parallel case in Finnish, and proposes a similar separation between the exception-filled sequence structure constraint and an exceptionless suffix harmony process. 
    \citet[171f.]{Howard1972} correctly observes that this duplication is also unnecessary.}

\citeauthor{SPE} assume that the specification of the target (i.e., the segment or segments to be changed) of a rule \emph{R} must be marked [$+$\emph{R}] by convention. 
A root or affix which fails to undergo \emph{R} despite otherwise matching the structural description is simply said to be marked [$-$\emph{R}]. 
In other words, no representation is ever truly an exception; rather, some underlying representations have non-default features which do not match the extended structural description of $R$, which requires that the target be [$+R$]. 
If disharmonic roots are marked [$-$\textsc{Backness Harmony}], then the final vowel of disharmonic roots will still trigger \textsc{Backness Harmony}, since the [$-$\textsc{Backness Harmony}] root is no longer the target but rather the trigger, which is not subject to the [$+$\textsc{Backness Harmony}] requirement.

Under this account, root and suffix harmony are derived from \textsc{Backness Harmony}, but they do not have the same ontology: suffix harmony is direct result of phonological rule application, whereas any ``effects'' of the  generalization concerning harmonic roots arises from a dispreference for lexical exceptionality.

\citeauthor{A74} also notes that roots which fail to undergo suffix harmony, like (\ref{turkexcept}b), may themselves be harmonic. He takes this to be evidence for the necessity of duplication:

@@

\noindent
Under the assumptions so far, there is no reason to think that \emph{kabahat} is [$+$\textsc{Backness Harmony}], however: even roots with all back vowels need not undergo this rule. Whatever the ultimate analysis of those few words which fail to show suffix harmony, nothing depends on whether or not the roots in question are harmonic.

It is necessary to dispense with an alternative analysis proposed by \citet{Clements1982} and \citet{Inkelas1997}. 
This analysis has desirable properties, but it subtlely reintroduces the duplication of phonological process and sequence structure constraint. 
Root vowels exhibit a robust contrast for backness (e.g., \emph{kül} `ash' vs. \emph{kul} `servant', \emph{kepek} `bran' vs. \emph{kapak} `lid'), whereas backness of vowels in non-initial syllables is predictable in harmonic roots (there are no prefixes in Turkish).
\citeauthor{Clements1982} propose that these vowels, as well as harmonizing suffix vowels, are underspecified for backness, whereas the non-initial vowels of disharmonic roots and of certain exceptional suffixes are fully specified. 
This is schematized below.

@@

One crucial detail is missing from this analysis: \textsc{Backness Harmony} needs to be prevented from overwriting the [$+$\textsc{Back}] specification of disharmonic roots, perhaps by structure preservation condition \citep{Kiparsky1985}. 
However, any condition which prevents \textsc{Backness Harmony} from overwriting underlying backness specifications will reintroduce the duplication of sequence structure and phonological generalizations; under this analysis, disharmonic roots are no longer exceptional, despite considerable evidence (reviewed below) that they are marked in Turkish.\footnote{
    On the other hand, it is possible to interpret the presence of a single backness specification per root as a sort of default. 
    A precedent for this is the surface-oriented interpretation of the tonal Obligatory Contour Principle proposed by \citet[134]{Goldsmith1976} and \citet{Odden1986}, under which adjacent identical tones are automatically attributed to a single underlying tone. 
    However, this is merely a notational variant of the rule exceptionality account in which [$+$\textsc{Backness Harmony}] is the default.} 
    The underspecification analysis is rejected here on these grounds.

While harmony in non-derived environments can be inferred from the aforementioned suffix alternations, no evidence has yet been presented to show that Turkish speakers internalize the tendency for roots to conform to backness harmony. 
    If Turkish speakers do not attend to this generalization, there is no need for the grammar to account for it. Several other ``external'' facts suggest that this is not the case. 
    The discussion here is not intended to imply uncritical acceptance of evidence from loanword adaptation, language games, or psycholinguistic tasks as evidence for phonological grammar, but rather to illustrate additional evidence that is pertinent if the linking hypothesis is correct.\footnote{
    Thanks to Bert Vaux and Kie Zuraw for bringing these studies to my attention.}

@@

Similar evidence comes from a language game discussed by \citet{Harrison2001}. 
    The game is native to the related language Tuvan, where it is used to convey a sense of ``vagueness or jocularity''; it is not indigenous to Turkish, but can be taught quickly to children or adults. 
    In this game, the base is reduplicated and the first vowel of the reduplicant replaced with a [$+$\textsc{Back}] vowel. 
    In (\ref{redupgame}a), the second [$-$\textsc{Back}] vowel of the base is, in the reduplicant, ``reharmonized'' with the inserted [$+$\textsc{Back}] vowel. 
    The disharmonic roots of (\ref{redupgame}b) do not reharmonize.\footnote{
    A similar contrast between harmonic and disharmonic roots is found in Tuvan \citep{Harrison2001} and in an unrelated Finnish language game \citep{Campbell1986}.}

@@

\noindent 
In the full specification analysis adopted here, reharmonization is the result of \textsc{Backness Harmony} applying within the reduplicant. 
    On the other hand, the lack of reharmonization in the reduplicants of disharmonic roots suggests that the [$-$\textsc{Backness Harmony}] exception feature is copied under reduplication.\footnote{
    There is reason to suspect that reduplicants bear the lexical diacritics of the roots they are derived from.
    In Kinande, verb reduplication, documented by \citet{Mutaka1990} and further discussed by \citet{Downing2000}, requires a bisyllabic reduplicant. 
    This clearly has synchronic force in the gramar, since reduplicated monosyllabic roots contain three copies of the root so as to satisfy this requirement, and reduplicated forms of many trisyllabic verbs are ineffable. 
    However, a few trisyllabic verbs exceptionally show full reduplication.
    This is a lexical property, but presence or absense of this exceptionalty feature is only apparent in the form of the reduplicant.}
    Under this definition, root harmony is not static, and so this is not an exception to the claim that static phonotactic generalizations are not preserved by operations like reduplication and truncation \citep{Silverman2000}.

@@

The Turkish word-spotting experiment is adapted for infants by \citet{Kampen2008}.
In this study, 9-month-old infants are familiarized with recordings of harmonic, disyllabic nonce words presented in isolation.
At test time, the infants listen to the disyllabic nonce words in isolation using the head turn preference paradigm. 
Infants acquiring Turkish listen longer to nonce words preceded by a disharmonic juncture during familiarization (e.g., \emph{lo-NETIS}), whereas infants acquiring German, a language which lacks vowel harmony, do not exhibit this preference. 
Similarly, \citeauthor{Kampen2008} report that Turkish 6-month-old infants prefer to listen to harmonic nonce words such as \emph{paroz} over disharmonic nonce words like \emph{nelok}, but German 6-month-old infants show no such preference.
However, there are some caveats for identifying these effects with grammatical computations: the domain for segmentation effects is considerably larger than the domain for harmony, so the two cannot be easily identified.
All that can be said is that there is an obvious similarity between computing the contexts for vowel harmony and for the word segmentation heuristic.

@@

If permitted to apply in non-derived environments, this rule accounts for the tendency of polysyllabic roots to contain only [$+$\textsc{Rnd}] or only [$-$\textsc{Rnd}] [$+$\textsc{High}] vowels.
In concert with \textsc{Backness Harmony}, \textsc{Roundness Harmony} also triggers alternations which account for the shape of the dative singular (dat.sg.) and genitive singular (gen.sg.) suffixes, among others.suffixes, among others.
As was the case for \textsc{Backness Harmony}, disharmonic roots are found, and the final vowel of disharmonic roots triggers suffix harmony.

@@

Counts were computed by regular expression matching on a 9,601-root subset of the TELL database which consists of roots which show no surface variation in any inflected form.

@@

>>>>

It has long been speculated that statistical criteria could distinguish between \emph{accidental phonotactic gaps}, those which could arise without any antecedent cause, and those gaps which are \emph{structural} in nature \citep[e.g.,][]{Fischer-Jorgensen1952,Saporta1955,Saporta1958,Vogt1954}.\footnote{
    Some of this early work is, however, critical of any attempt to establish a firm distinction between structural and accidental gap.
    For instance, \citet[3]{Fischer-Jorgensen1952} writes that ``it is theoretically impossible to fix a non-arbitrary borderline between law and accident.''}
Following seminal work by \citet{Mester1988} and \citet{McCarthy1988}, it is now commonly assumed that the phonotactic patterns in lexical entries ``directly determine the mental representation of the phonotactic constraints'' \citep[180]{Frisch2004} acquired by speakers, and therefore phonotactic constraints, whether categorical or gradient, can be directly inferred from statistical analysis of the lexicon.
There is by now an enormous amount of research that adopts this assumption.
Afroasiatic languages, especially Arabic \citep{Coetzee2008a,Frisch2004,McCarthy1988,Pierrehumbert1993} but also Berber \citep{Elmedlaoui1995} and Tigrinya \citep{Buckley1997}, and Austronesian languages, including Javanese \citep{GraffInPress,Mester1988}, Muna \citep{Anttila2008a,Coetzee2008a}, and Samoan \citep{AldereteInPress}, have been of particular interest to linguists adopting this hypothesis; Aymara \citep{GraffInPress}, Cantonese \citep{Yip1989}, English \citep{Berkley1994a,Berkley1994b,Berkley2000,Coetzee2008b,Davis1989,Martin2007,Martin2011}, Gitksan \citep{Brown2010}, Hungarian \citep{Grimes2010}, Japanese \citep{Kawahara2006}, Ju|'hoansi \citep{Kinney2005}, Navajo \citep{Martin2007,Martin2011}, Ofo \citep[38f.]{MacEachern1999}, Russian \citep{Padgett1991,Padgett1992}, Shona \citep{Hayes2008a}, and Wargamay \citep{Hayes2008a} have also received a statistical treatment. 
\citet{Pozdniakov2007} analyze lexical tendencies in a diverse sample of 30 languages.

@@

It would be a result of great interest were it shown that statistical significance is both necessary and sufficient to identify linguistic generalizations which are internalized by speakers, but there is no reason to grant this assumption with respect to phonotactic knowledge.
Nothing demands that the antecedent cause of a statistically reliable lexical tendency be grammatical. 
On the contrary: there are reasons to suspect that many static phonotactic constraints identified in this manner have no synchronic reality.

@@

\noindent
However, subsequent sound changes \citep[e.g.,][]{Baldi1994,Safarewicz1932}, particularly the degemination of Old Latin \emph{ss} after diphthongs and long monophthongs (e.g., \emph{caussa} $>$ \emph{causa} `cause'), introduce numerous exceptions to \textsc{Rhotacism}.
In Classical Latin, intervocalic \emph{s} is found root-internally (\emph{asellus} `donkey', \emph{casa} `hut'), in environments derived by inflectional suffixes (\emph{uāsis} `vase' \textsc{gen.sg.}, \emph{uisēre} `to view'), prefixation (\emph{dēsecāre} `to cut off'; cf.~\emph{dē} `from', \emph{secāre} `to cut'), compounding (\emph{olusātrum} `parsnip'; cf.~\emph{olus} `vegetable', \emph{ātrum} `black'), and denominal adjective formation (\emph{uentōsus} `windy'; cf.~\emph{uentus} `wind'), and is tolerated in nativized loanwords from Celtic (\emph{omāsum} `tripe'), Germanic (\emph{glaesum} `amber'), and Greek (\emph{basis} `pedestal').
These facts lead Saussure to conclude that \textsc{Rhotacism} is no longer ``inhérente à la nature de la langue'' (202).
While some of the apparent exceptions may be the result of opaque phonological application \citep{Heslin1987}---an explanation not yet available in Saussure's time---any formulation of rhotacization will admit nearly as many lexical exceptions as there are roots exhibiting \emph{s}-\emph{r} alternations \citep{GormanInPressc}.
Any synchronic account of the underrepresentation of intervocalic \emph{s} must confront the unproductive nature (as indicated by the accumulation of exceptions) of the proximate explanation for this tendency.

In other cases, it is possible to observe once robust phonotactic tendencies induced by sound change become increasingly moribund in real time.
Perhaps the most famous example concerns the distribution of short \emph{a} in English.
Early Modern English short \emph{a} underwent irregular lengthening and tensing before voiceless fricatives /f, θ, s/ and nasals /m, n/ \citep[I.203f.]{AOE}, introducing the first traces of the complex pattern now found in many American dialects (as well as in various British dialects, e.g., \citealp[74f.]{Jones1964}).
A great deal of variation in Mid-Atlantic speech can be summarized as follows: the low front vowel is lax except before tautosyllabic /f, θ, s, m, n/ (and perhaps other consonants), where the low front vowel is tense (and perhaps raised), with various exceptions in both directions \citep[e.g.,][]{Cohen1970,Ferguson1975,Labov1981,Trager1930,Trager1934,Trager1940}.
\citet{Cohen1970}, \citet{Labov1981}, and \citet{Trager1940} all argue that this tendency is no longer to be attributed to synchronic allophony, but to a largely-complete lexical/phonemic split.
\citeauthor{Labov1981} presents three arguments for this analysis.
First, there are minimal pairs (tense modal \emph{can} vs.~lax noun/verb \emph{can}) and extensive individual variation.
Secondly, \citet{Payne1980} finds that children born out-of-state are unable to acquire the short \emph{a} pattern of the Philadelphia metropolitian area in later childhood, though they participate fully in other sound changes specific to the area.
Finally, Mid-Atlantic speakers have a greater ability to discriminate tense and lax variants of short \emph{a} than speakers of dialects with allophonic short \emph{a} tensing/laxing.
In summary, the historical facts that account for this distribution no longer play a clear role in the synchronic phonology.

Other cases show that many constraints identified by statistical analysis are inert: that is, there is no evidence that they are internalized by speakers.
For instance, most instances of Modern English [ʃ] derive from Old English [sk] (e.g., \emph{fisc} `fish') via sound change.
Since Old English does not permit long vowels before complex codas, compared to similar segments like [s], [ʃ] is still rarely preceded by long vowels in word-final syllables (Table \ref{ssh}).\footnote{
    This sample is drawn from the CMU pronunciation dictionary, filtered by excluding words with a token frequency of less than 1 per million words in the SUBTLEX-US frequency norms; similar results can be obtained with less restrictive samples.
    The four words ending in a long vowel-[ʃ] sequence are \emph{douche}, \emph{leash}, \emph{unleash}, and \emph{woosh}.}
As can be seen, long vowels are twice as common before [s] as before [ʃ], a significant generalization according to the Fisher exact test.
Similarly, \citet{HayesInPress} report that this constraint is discovered by the \citet{Hayes2008a} phonotactic learning model, which uses a related statistical criterion to identify constraints.
Despite this, \citet{Iverson2005} label the constraint on long vowels before [ʃ] as ``phonologically accidental'', as a millennium of coinages (e.g., 
`affective'' \emph{woosh}) and loanwords (e.g., \emph{douche}) disregard this generalization.
\citet{HayesInPress} find that a variant of this restriction has little or no effect on wordlikeness judgements.

\begin{table}[t]
\centering
\begin{tabular}{l r r r r}
\toprule
          & \{ɪ, ʊ\}\gap{}\# & \{i, u\}\gap{}\# & \% long & $p$-value \\
\midrule
\gap{}ʃ\# & 46               & 4                & 8\%     & \multirow{2}{*}{1.67\e{-06}} \\
\gap{}s\# & 40               & 36               & 53\%    & \\
\bottomrule
\end{tabular}
\caption{Type frequencies of high vowels before word-final [ʃ] and [s]}
\label{ssh}
\end{table}

This latter example makes it clear that a purely statistical criterion overgenerates in the sense that it predicts phonotactic constraints which speakers do not seem to internalize.
To account for the cases like the one above, \citet{HayesInPress} propose that speakers are biased in favor of ``natural generalizations'' in probabilistic phonotactic learning.\footnote{
    \citeauthor{HayesInPress} do not provide an operational definition of ``naturalness'', so it is difficult to evaluate their specific results despite their relevance to the argument at hand.}
However, this chapter argues that this overgeneration gives the lie to the broader assumption that phonotactics are extracted from patterns in the lexicon.
Rather, the only restrictions which speakers internalize are those which derive from phonological processes in the language.
This is independent of ``naturalness'', since both ``natural'' and ``unnatural'' variants of statistically reliable static constraints are shown to be inert.

@@

\citet{Lees1966b,Lees1966a} proposes three constraints on Turkish vowel sequences; these constraints are the focus of many subsequent studies. In this section, these constraints are formalized, and where possible, related to phonological alternations and to behavioral evidence bearing on speakers' knowlege of the restrictions. The following feature specification for the eight vowels of Turkish is assumed throughout \citep[e.g.,][298]{Dresher2009}.\footnote{\citeauthor{Dresher2009} in fact deploys a feature labeled \textsc{Labial} rather than \textsc{Rnd}.}

@@

Two less common notations are used in this chapter.
First, directional application conditions are assumed, so as to derive the left-to-right, iterative properties of the harmony rules.
Since it was first proposed by \citet{Johnson1972}, considerable evidence for directional application has been adduced (e.g., \citealt{A74}: chap.~9, \citealt{GormanInPressc}, \citealt{Howard1972}:65f., \citealt{Kavitskaya2008}, \citealt{Kaye1982}, \citealt{KK77}:189f., \citeyear{KK79}:319f., \citealt{Piggott1975}, \citealt{Sohn1971}; see \citealt{McCarthy2003b} and \citealt{Wolf2011b} for recent reviews.)
Further, \citet{Johnson1972} and \citet{Kaplan1994} prove directional application and simultaneous application are equivalent in terms of formal learnability.

Secondly, rather than the use of an unbounded number of Greek-letter variables ($\alpha$, $\beta$, etc.) over feature values $\{+, -\}$, only a single variable, denoted by `$=$', is used \citep{McCawley1973}. 
A structural description [$=$F]\ldots{}[$=$F] matches a string S$_i$\ldots{}S$_j$ if and only if S$_i$ and S$_j$ are both [$+$F] or both [$-$F].
This is more restrictive than Greek-letter notation, in that it prevents the value of one feature being applied to different feature.\footnote{
    \citet{Odden2012} argues that the two counterexamples against this restriction presented in \emph{SPE} (352-353) are not probative.}

@@

\citet[212]{A74} and \citet{Iverson1978} argue that suffix harmony in disharmonic roots found in (\ref{turkexcept}a) requires the rule governing suffix harmony alternations to be distinguished from a sequence structure constraint governing root harmony. 
Since the rule and sequence structure constraint are otherwise identical, this constitutes a ``duplication'' (in the sense of \citealt{KK77}:136f.): harmony is stated at two points in the grammar.

Another possiblity, however, is to understand root disharmony as a type of lexical exceptionality.
Under such an analysis, suffix harmony in disharmonic roots is entirely consistent with the theory of exceptionality proposed in \emph{SPE} \citep[197f.]{Zonneveld1978}.\footnote{
    \citet[29f.]{Kiparsky1968} discusses a parallel case in Finnish, and proposes a similar separation between the exception-filled sequence structure constraint and an exceptionless suffix harmony process.
    \citet[171f.]{Howard1972} suggests an analysis of Finnish that is quite similar to the one proposed for Turkish below.}
\citeauthor{SPE} assume that the specification of the target (i.e., the segment or segments to be changed) of a rule \emph{R} must be marked [$+$\emph{R}] by convention.
A root or affix which fails to undergo \emph{R} despite otherwise matching the structural description is simply said to be marked [$-$\emph{R}]. 
In other words, no representation is ever truly an exception to a rule; rather, some underlying representations have non-default features which do not match the extended structural description of $R$, which requires that the target be [$+R$].
If disharmonic roots are marked [$-$\textsc{Backness Harmony}], then the final vowel of disharmonic roots will still trigger \textsc{Backness Harmony}, since the [$-$\textsc{Backness Harmony}] root is no longer the target but rather the trigger, which is not subject to the [$+$\textsc{Backness Harmony}] requirement.
Under this account, root and suffix harmony are both derived from \textsc{Backness Harmony}, but they do not have the same ontology: suffix harmony is direct result of phonological rule application, whereas the tendency for harmonic roots arises from a dispreference for lexical exceptionality.
Under this analysis, there is no need to state a sequence structure constraint producing root harmony.

\citeauthor{A74} also notes that a small number of roots which fail to undergo suffix harmony, like (\ref{turkexcept}b), may themselves be harmonic.
He takes this to be evidence for the necessity of duplication:

@@

\noindent
Under the assumptions so far, however, there is no reason to assume that \emph{kabahat} (or any other root which fails to trigger suffix harmony) is [$+$\textsc{Backness Harmony}].
The tendency for root harmony holds of the Turkish lexicon as a whole, but there is no evidence that it holds of those few roots which select exceptional suffixes; indeed, there are disharmonic roots which select disharmonic suffixes (e.g., \emph{etol} `fur stole', nom.pl.~\emph{etoller}).

This analysis compares favorably to an alternative proposed by \citet{Clements1982} and \citet{Inkelas1997} which has other desirable properties but
which makes no predictions about root harmony.
Root vowels exhibit a robust contrast for backness (e.g., \emph{kül} `ash' vs.~\emph{kul} `servant', \emph{kepek} `bran' vs.~\emph{kapak} `lid'), whereas backness of vowels in non-initial syllables is predictable in harmonic roots (note that there are no prefixes in Turkish).
\citeauthor{Clements1982} and \citeauthor{Inkelas1997} propose that these vowels, as well as harmonizing suffix vowels, are underspecified for backness, whereas the non-initial vowels of disharmonic roots and of certain exceptional suffixes are fully specified.
This is schematized below.

@@

One crucial detail is missing from this analysis: \textsc{Backness Harmony} needs to be prevented from overwriting the [$+$\textsc{Back}] specification of disharmonic roots, perhaps by structure preservation condition \citep{Kiparsky1985}. 
However, such a condition makes it impossible to relate \textsc{Backness Harmony} to the tendency for harmonic roots, despite considerable evidence (reviewed below) that they are marked in Turkish.\footnote{
    On the other hand, it is possible to interpret the presence of a single backness specification per root as a sort of default.
    A precedent for this is the surface-oriented interpretation of the tonal Obligatory Contour Principle proposed by \citet[134]{Goldsmith1976} and \citet{Odden1986}, under which adjacent identical tones are automatically attributed to a single underlying tone. 
    However, this is merely a notational variant of the rule exceptionality account in which [$+$\textsc{Backness Harmony}] is the default.}

While harmony in non-derived environments can be inferred from the aforementioned suffix alternations, no evidence has yet been presented to show that Turkish speakers internalize the tendency for roots to conform to backness harmony. 
    If Turkish speakers do not attend to this generalization, there is no need for the grammar to account for it. Several other ``external'' facts suggest that this is not the case. 
    The discussion here is not intended to imply uncritical acceptance of evidence from loanword adaptation, language games, or particular psycholinguistic tasks as evidence for phonological grammar, but rather to illustrate additional evidence that is pertinent if the linking hypothesis is correct.\footnote{
    Thanks to Bert Vaux and Kie Zuraw for bringing these studies to my attention.}

@@

Similar evidence comes from a language game discussed by \citet{Harrison2001}. 
    The game is native to the related language Tuvan, where it is used to convey a sense of ``vagueness or jocularity''; it is not indigenous to Turkish, but can be taught quickly to children or adults. 
    In this game, the base is reduplicated and the first vowel of the reduplicant replaced with a [$+$\textsc{Back}] vowel. 
    In (\ref{redupgame}a), the second [$-$\textsc{Back}] vowel of the base is, in the reduplicant, ``reharmonized'' with the inserted [$+$\textsc{Back}] vowel. 
    The disharmonic roots of (\ref{redupgame}b) do not reharmonize.\footnote{
    A similar contrast between harmonic and disharmonic roots is found in Tuvan \citep{Harrison2001} and in an unrelated language game in Finnish \citep{Campbell1986}.}

@@

\noindent
\citeauthor{Harrison2001} propose that those roots which fail to reharmonize are prespecified for backness throughout; this also requires that reharmonization is subject to a non-derived environment condition.
In the full specification analysis adopted here, reharmonization is the result of \textsc{Backness Harmony} applying within the reduplicant; the lack of reharmonization in the reduplicants of disharmonic roots requires that the [$-$\textsc{Backness Harmony}] exception feature is copied under reduplication.\footnote{
    There are other cases which suggest that lexical diacritics are copied in reduplication.
    In Kinande, verbal reduplication requires a bisyllabic reduplicant \citep{Downing2000,Mutaka1990}.
    This requirement has synchronic force in the grammar, since ``reduplicated'' monosyllabic roots in fact contain three copies of the root (so that the reduplicant is bisyllabic); furthermore, reduplicated forms of many trisyllabic verbs are ineffable.
    However, a few trisyllabic verbs exceptionally show full reduplication.
    This is a lexical property, but presence or absence of the exceptionalty feature that permits non-bisyllabic reduplicants only surfaces on the reduplicant itself.}

@@

The Turkish word-spotting experiment is adapted for infants by \citet{Kampen2008}.
In this study, 9-month-old infants are familiarized with recordings of harmonic, disyllabic nonce words presented in isolation and tested with the head turn preference paradigm.
Infants acquiring Turkish listen longer to nonce words preceded by a disharmonic juncture during familiarization (e.g., \emph{lo-NETIS}), whereas infants acquiring German, a language which lacks vowel harmony, do not exhibit this preference. 
Similarly, \citeauthor{Kampen2008} report that Turkish 6-month-old infants prefer to listen to harmonic nonce words such as \emph{paroz} over disharmonic nonce words like \emph{nelok}, but German 6-month-old infants show no such preference.
However, there are some caveats for identifying these effects with grammatical computations: the domain for segmentation effects is considerably larger than the domain for harmony, so the two cannot be easily identified.
All that can be said is that there is an obvious similarity between computing the contexts for vowel harmony and for the word segmentation heuristic.

@@

If permitted to apply in non-derived environments, this rule accounts for the tendency of polysyllabic roots to contain only round or unround high vowels.
In concert with \textsc{Backness Harmony}, \textsc{Roundness Harmony} also triggers alternations which account for the shape of the dative singular (dat.sg.) and genitive singular (gen.sg.) suffixes, among others.suffixes, among others.
As was the case for \textsc{Backness Harmony}, disharmonic roots occur, but generally trigger suffix harmony.

@@

Counts are computed by regular expression matching on a 9,601-root subset of the TELL database, consisting of roots which show no surface variation in any inflected form.

@@

Speakers do not have the preferences predicted by \textsc{Labial Attraction}.
At the same time, they do not have a preference for \textsc{Roundness Harmony} either, as one might expect.
At most, it could be said that \textsc{Labial Attraction} is sufficiently robust so as to suspend the effect of \textsc{Roundness Harmony}, with which it is in competition.
